Evolution Reinforces Cooperation with the Emergence of Self-Recognition Mechanisms: an Empirical Study of the Moran process for the Iterated Prisoner's Dilemma

### 2017-10-31 main.tex
Add better explanation of the cooperation rates.
Also remove the names from the plot.

% Title
\title{Evolution Reinforces Cooperation with the Emergence of Self-Recognition
       Mechanisms: an Empirical Study of the Moran process for the Iterated
       Prisoner's Dilemma}
\author[1]{Vincent Knight}
\author[2]{Marc Harper}
\author[1]{Nikoleta E. Glynatsi}
\author[3]{Owen Campbell}

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic} for Alternator and WSLS.
The top most line on all figures (in red and using a circle) corresponds to
\(x_{N-1}\), the middle line (in green and using a cross) corresponds to
\(x_{N/2}\) and the bottom line (in blue and using an x) corresponds to
\(X_{1}\).
The points represent the simulated
values and the line shows the theoretical value. Note that these are 
deterministic strategies and show a good match between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategy pairs. These means have been compared using a \(t\)-test and the \(p\)
values are shown in Table~\ref{tab:comparison_deterministic} which confirms the
fact that the theoretic and simulated values are a good match.

Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies: Calculator and arrogant Q Learner. These are no longer a
good match (confirmed with a \(t\)-test in
Table~\ref{tab:comparison_stochastic}) which highlights the weakness of assuming
a given interaction between two IPD strategies can be summarised with a set of
utilities as shown in (\ref{equ:payoff_matrix}). For any given pair of
strategies it is possible to obtain \(p_{i,i-1}, p_{i,i+1}, p_{ii}\) exactly (as
opposed to the approximations offered by (\ref{equ:p_up}), (\ref{equ:p_down})
and (\ref{equ:p_stay})). Obtaining these requires particular analysis for a
given pair and can be quite a complex endeavour for stochastic strategies with
long memory: this is not necessary for the purposes of this work.  All data
generated for this validation exercise can be found at~\cite{data}.

\begin{figure}[!hbtp]
    \centering
    \includegraphics[width=.95\columnwidth]{img/Alternator_v_WSLS.pdf}
    \caption{Comparison of theoretic and actual Moran Process fixation
             probabilities for \textbf{deterministic} strategies: Alternator and
             Cooperator.}
    \label{fig:comparison_deterministic}
\end{figure}

\begin{table}[!hbtp]
    \centering
    \input{tex/t_test_results_alternator_v_wsls.tex}
    \caption{\(p\) values resulting from a $t$ test comparing the theoretic
        value with the simulated value of the Moran Process fixation
        probabilities for \textbf{deterministic} strategies: Alternator and
    Cooperator.}
    \label{tab:comparison_deterministic}
\end{table}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[width=.95\columnwidth]{img/Calculator_v_Arrogant_QLearner.pdf}
    \caption{Comparison of theoretic and actual Moran Process
             fixation probabilities for \textbf{stochastic} strategies:
             Calculator and Arrogant Q Learner.}
    \label{fig:comparison_stochastic}
\end{figure}

\begin{table}[!hbtp]
    \centering
    \input{tex/t_test_results_calculator_v_arrogant_q_learner.tex}
    \caption{\(p\) values resulting from a \(t\) test comparing the theoretic
        value with the simulated value of the Moran Process fixation
        probabilities for \textbf{stochastic} strategies: Calculator and
    Arrogant Q Learner.}
    \label{tab:comparison_stochastic}
\end{table}

@@

When $N=2$ the Moran process is effectively a measure of the distribution of relative
mean payoffs over all possible matches between two players. The strategy
that scores higher than the other more often will fixate more often. For \(N=2\)
the two cases of \(x_1\) and \(x_{N-1}\) coincide, but will be
considered separately for larger \(N\) in sections~\ref{sec:strong_invaders}
and~\ref{sec:strong_resistors}. Figure~\ref{fig:boxplot_2} shows all fixation
probabilities for the strategies considered. The top 16 (10\%) strategies are
% TODO Fix (have removed plot)
shown in
Table~\ref{tbl:summary_top_2}. The top five ranking strategies are:

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism described above.
    \item Defector: it always defects. Since it has no interactions with other
        defectors (recall that \(N=2\)), its aggressiveness is rewarded.
    \item Aggravater, which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in \cite{Ashlock2006}.
    \item Handshake, a slightly less aggressive version of the Collective
        Strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself just as the Collective Strategy does. The
        difference is that CS will defect after the handshake if the opponent
        defects while Handshake will not.
\end{enumerate}

@@

\begin{table}[!hbtp]
    \tiny
    \centering
    \scalebox{0.95}{\input{tex/summary_top_2_invade.tex}}
    \caption{Top strategies for \(N=2\) (neutral fixation is \(p=0.5\))}
    \label{tbl:summary_top_2}
    % TODO Explain this more (see comment about error bars)
\end{table}

@@

@@

@@

\begin{table}[!hbtp]
    \centering
    \tiny
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \input{tex/summary_top_3_invade.tex}
        \caption{\(N=3\)}
    \end{subfigure}%

@@

@@

@@

\begin{table}[!hbtp]
    \tiny
    \centering
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \input{tex/summary_top_3_resist.tex}
        \caption{\(N=3\)}
    \end{subfigure}%

@@

@@

\begin{figure}[!htbp]
    \centering
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/correlation_heatmap_invade.pdf}
        \caption{Rank based on \(x_1\)}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/correlation_heatmap_resist.pdf}
        \caption{Rank based on \(x_{N - 1}\)}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/correlation_heatmap_coexist.pdf}
        \caption{Rank based on \(x_{N/2}\)}
    \end{subfigure}
    \caption{Heatmap of correlation coefficients of rankings by population size.}
    \label{fig:correlation_coefficients}
\end{figure}

@@

\section{List of players}\label{app:list_of_players}

\footnotesize
\begin{enumerate}
    \input{tex/list_of_players.tex}
\end{enumerate}

>>>>

% Title
\title{Evolution Reinforces Cooperation with the Emergence of Self-Recognition
       Mechanisms: an Empirical Study of the Moran process for the Iterated
       Prisoner's Dilemma}
       % TODO Possibly change
\author[1]{Vincent Knight}
\author[2]{Marc Harper}
\author[1]{Nikoleta E. Glynatsi}
\author[3]{Owen Campbell}

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic}. The points represent the simulated
% TODO Add description of lines shown.
values and the line shows the theoretical value. Note that these are all
deterministic strategies and show a perfect match between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategy pairs. Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
% TODO Rephrase this sentence: possibly when doing the statistical test
weakness of assuming a given interaction between two IPD strategies can be
summarised with a set of utilities as shown in
(\ref{equ:payoff_matrix}). For any given pair of strategies it is possible to
obtain \(p_{i,i-1}, p_{i,i+1}, p_{ii}\) exactly (as opposed to the
approximations offered by (\ref{equ:p_up}), (\ref{equ:p_down}) and
(\ref{equ:p_stay})). Obtaining these requires particular analysis for a given
pair and can be quite a complex endeavour for stochastic strategies with long
memory: this is not necessary for the purposes of this work.
All data generated for this validation exercise can be found
at \cite{data}.
% TODO Include a statistical test (Bonferroni)

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Alternator_v_Cooperator.pdf}
        \caption{Alternator and Cooperator}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Defector_v_Alternator.pdf}
        \caption{Defector and Alternator}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Alternator_v_TfT.pdf}
        \caption{Alternator and Tit For Tat}
    \end{subfigure}%

    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Cooperator_v_TfT.pdf}
        \caption{Cooperator and Tit For Tat}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Defector_v_Cooperator.pdf}
        \caption{Defector and Cooperator}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Defector_v_TfT.pdf}
        \caption{Defector and Tit For Tat}
    \end{subfigure}%

    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/WSLS_v_TfT.pdf}
        \caption{Win Stay Lose Shift and Tit For Tat}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Alternator_v_WSLS.pdf}
        \caption{Alternator and Win Stay Lose Shift}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Defector_v_WSLS.pdf}
        \caption{Defector and Win Stay Lose Shift}
    \end{subfigure}%
    \caption{Comparison of theoretic and actual Moran Process fixation
             probabilities for \textbf{deterministic} strategies.}
    \label{fig:comparison_deterministic}
    % TODO DEEMED Unreadable: reduce number of images.
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Random_v_Cooperator.pdf}
        \caption{Random and Cooperator}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Random_v_Defector.pdf}
        \caption{Random and Defector}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Random_v_TfT.pdf}
        \caption{Random and Tit For Tat}
    \end{subfigure}%

    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/ALLCorALLD_v_Cooperator.pdf}
        \caption{All C or all D and Cooperator}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/ALLCorALLD_v_Defector.pdf}
        \caption{All C or all D and Defector}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Calculator_v_Random.pdf}
        \caption{Calculator and Random}
    \end{subfigure}%

    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Calculator_v_ALLCorALLD.pdf}
        \caption{Calculator and All C or all D}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/Calculator_v_Arrogant_QLearner.pdf}
        \caption{Calculator and Arrogant Q learner}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.8\columnwidth]{img/ALLCorALLD_v_TfT.pdf}
        \caption{All C or all D and Tit For Tat}
    \end{subfigure}%
    \caption{Comparison of theoretic and actual Moran Process
             fixation probabilities for \textbf{stochastic} strategies.}
    \label{fig:comparison_stochastic}
    % TODO DEEMED Unreadable: reduce number of images.
\end{figure}

@@

When $N=2$ the Moran process is effectively a measure of the distribution of relative
mean payoffs over all possible matches between two players. The strategy
that scores higher than the other more often will fixate more often. For \(N=2\)
the two cases of \(x_1\) and \(x_{N-1}\) coincide, but will be
considered separately for larger \(N\) in sections~\ref{sec:strong_invaders}
and~\ref{sec:strong_resistors}. Figure~\ref{fig:boxplot_2} shows all fixation
probabilities for the strategies considered. The top 16 (10\%) strategies are
shown in
Table~\ref{tbl:summary_top_2}. The top five ranking strategies are:

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism described above.
    \item Defector: it always defects. Since it has no interactions with other
        defectors (recall that \(N=2\)), its aggressiveness is rewarded.
    \item Aggravater, which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in \cite{Ashlock2006}.
    \item Handshake, a slightly less aggressive version of the Collective
        Strategy \cite{Robson1990}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself just as the Collective Strategy does. The
        difference is that CS will defect after the handshake if the opponent
        defects while Handshake will not.
\end{enumerate}

@@

\begin{figure}[!hbtp]
    \begin{subfigure}{.5\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/boxplot_2_invade.pdf}
        \caption{The fixation probabilities for \(N=2\)}
        \label{fig:boxplot_2}
    \end{subfigure}%
    ~
    \begin{subfigure}{.5\columnwidth}
        \scriptsize
        \centering
        \scalebox{0.95}{\input{tex/summary_top_2_invade.tex}}
        \caption{Top strategies for \(N=2\) (neutral fixation is \(p=0.5\))}
        % TODO Add more summary statistics to this measure
        \label{tbl:summary_top_2}
    \end{subfigure}
    \caption{Performance of strategies for \(N=2\).}
    % TODO Move the plot
    % TODO Explain this more (see comment about error bars)
\end{figure}

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[width=\columnwidth]{img/average_rank_vs_population_size_invade.pdf}
    \caption{\textbf{Invasion}: Ranks of all strategies according to \(x_1\) for different
    population sizes.}
    \label{fig:ranks_v_size_invade}
    % TODO Move to online appendix
\end{figure}

@@

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}{.3\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/boxplot_3_invade.pdf}
        \caption{\(N=3\)}
        \label{fig:boxplot_3_invade}
    \end{subfigure}%
    ~
    \begin{subfigure}{.3\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/boxplot_7_invade.pdf}
        \caption{\(N=7\)}
        \label{fig:boxplot_7_invade}
    \end{subfigure}%
    ~
    \begin{subfigure}{.3\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/boxplot_14_invade.pdf}
        \caption{\(N=14\)}
        \label{fig:boxplot_14_invade}
    \end{subfigure}
    \caption{The fixation probabilities \(x_1\).}
    % TODO Move to online appendix
\end{figure}

@@

\begin{table}[!hbtp]
    \centering
    \scriptsize
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \input{tex/summary_top_3_invade.tex}
        \caption{\(N=3\)}
    \end{subfigure}%

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[width=\columnwidth]{img/average_rank_vs_population_size_resist.pdf}
    \caption{\textbf{Resistance}: Ranks of all strategies according to \(x_{N-1}\) for different
    population sizes.}
    \label{fig:ranks_v_size_resist}
\end{figure}

@@

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}{.3\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/boxplot_3_resist.pdf}
        \caption{\(N=3\)}
        \label{fig:boxplot_3_resist}
    \end{subfigure}%
    ~
    \begin{subfigure}{.3\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/boxplot_7_resist.pdf}
        \caption{\(N=7\)}
        \label{fig:boxplot_7_resist}
    \end{subfigure}%
    ~
    \begin{subfigure}{.3\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{img/boxplot_14_resist.pdf}
        \caption{\(N=14\)}
        \label{fig:boxplot_14_resist}
    \end{subfigure}
    \caption{The fixation probabilities \(x_{N-1}\).}
    % TODO Move to online appendix
\end{figure}

@@

\begin{table}[!hbtp]
    \scriptsize
    \centering
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \input{tex/summary_top_3_resist.tex}
        \caption{\(N=3\)}
    \end{subfigure}%

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[width=\columnwidth]{img/average_rank_vs_population_size_coexist.pdf}
    \caption{Fixation ranks of all strategies according to \(x_{N/2}\) for different
    population sizes.}
    \label{fig:ranks_v_size_coexist}
    % TODO Move to online appendix
\end{figure}

@@

\begin{figure}[!htbp]
    \centering
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.9\columnwidth]{img/correlation_heatmap_invade.pdf}
        \caption{Rank based on \(x_1\)}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.9\columnwidth]{img/correlation_heatmap_resist.pdf}
        \caption{Rank based on \(x_{N - 1}\)}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{.3\columnwidth}
        \centering
        \includegraphics[width=.9\columnwidth]{img/correlation_heatmap_coexist.pdf}
        \caption{Rank based on \(x_{N/2}\)}
    \end{subfigure}
    \caption{Heatmap of correlation coefficients of rankings by population size.}
    % TODO Increase font
    \label{fig:correlation_coefficients}
\end{figure}

@@

% TODO Reduce font
% TODO Add number of states for FSMs
\section{List of players}\label{app:list_of_players}

\begin{enumerate}
    \input{tex/list_of_players.tex}
\end{enumerate}
